On the Functionality of the Multilayer Perceptron

### 2020-10-02 Thesis/Drafts/Buell-PHYS799-ThesisDraft0.tex
Updates to Thesis draft 0
- Cleaned up "Feature Selection" / "Spectrogram" Subsection

\paragraph*{}Tuomas Virtanen, machine learning and audio engineer writes in his book, "Computational Analysis of Sound Scene and Events" \cite{Virtanen}:
\begin{quote}
For recognition algorithms, the necessary property of the acoustic features is low variability among features extracted from examples assigned to the same class, and at the same time high variability allowing distinction between features extracted from examples assigned to different classes.
\end{quote}
In constructing a neural network classifier, the development of appropriate features is of the utmost importance. To ensure the construction of a suitable model, we derive features based from three sources (i) a spectrogram matrix of the waveform, (ii) the time-space representation of the waveform, and (iii) the frequency space representation of the waveform. It is important to note that although this algorithm will classify sound waves to instruments, the model will never actually be presented with a waveform directly, instead it will rely on these features.

\paragraph*{}Once we produce an sufficient set of features, we concatenate them into a single object, $\hat{x}$, called the \textit{feature-vector} \cite{Goodfellow}. In the training process, this object, along with the appropriate classification label, $y$ is presented to the neural network for processing. This process of constructing a feature vector from any data set is vital and is used to represent the data set in a far more compact and non-redundant format \cite{Virtanen,Liu}

@@

\paragraph*{}The $21$ features are from each file are organized into one of two array structures. each called a \textit{design matrix}. A design matrix is an N-Dimensional array of real floating-point numbers that stores the features in a convenient way for the neural network to digest. For this hybrid neural-network, we use two matrices, $X_1$ and $X_2$, each containing data from $b$ audio file samples. each has shape given by:
\begin{equation}
\label{eqn-X1Shape}
X_1 \in \mathbb{R}^{(b \times N' \times k' \times 1)}
\end{equation}
\begin{equation}
\label{eqn-X2Shape}
X_2 \in \mathbb{R}^{(b \times p)}
\end{equation}
Where $N' \times k' \times 1$ gives the shape of a single spectrogram matrix $S$, described in sect. (\ref{subsec-spectrogram}), The first axis, with length $b$ indicates that there are $b$ samples stored in that design matrix.
Similarly, $p$ is the number of features derived from the time and frequency-representation of features outlined in sec. (\ref{subsec-time}) and sec.(\ref{subsec-frequency}). A row of $X_2$ has shape $1 \times p$ where there are $p$ features within the \textit{feature-vector} for each sample. Matrix $X_1$ is presented to the convolutional branch of the network to be process as an a collection of images, and matrix $X_2$ is presented to the perceptron branch to be processed as a collection of vectors.

@@

\paragraph*{}The field of neural classification is well studied in the application of image-processing. Many large-scale, and even introductory neural network projects find themselves under the umbrella of image classification \cite{Geron,Goodfellow,Loy,Mierswa}. As a result, model architectures for image processing related tasks are well-explored and have shown experimentally successful behavior. Following this success, it make senses to provide an image-like representation of a sound wave as a feature. We do this in the form of a spectrogram matrix.

\paragraph*{}A spectrogram is a representation of the energy distribution of a sound wave as a function of both space and time. In a conventional spectrogram, the passing of time is shown along the $x$-axis, and the frequency spectrum is shown on the $y$-axis. Thus each point in the 2-Dimensional space is an energy at a given time and frequency. Examples spectrograms from the wave form data set are shown in Fig. (\ref{fig-spectrograms}).

@@

\paragraph*{}A spectrogram is produced by the method of \textit{frame-blocking}, which is very prevalent in audio signal classification. Frame-blocking takes a raw waveform or signal, $s$ and decomposes it into a set of analysis frames, $a_i$, with each being $N$ samples in length, and has a fixed overlap with the next adjacent frame. Each of the $k$ frames then allows for a section of the signal in somewhat stationary state \cite{Liu,Zhang,Kahn,Virtanen}. For this project, we have chosen to use frames of size $N = 4096$ with a $75\%$ or $3072$ sample overlap. At a sample rate of $f_s = 44100$ samples/second, each frame represents a slice of time that is about $0.1$ seconds long.

@@

\paragraph*{}After frame-blocking, we apply a \textit{windowing function} to each frame. A standard \textit{Hann Window} of $N$ samples is generated and reshaped into a $N \times 1$ column-array, $H$. The $n$-th index in a Hann window with $N$ samples is defined:
\begin{equation}
\label{eqn-Hann}
H[n] = \frac{1}{2}\bigg[ 1 - \cos\Big( \frac{2\pi n}{N-1}\Big)\bigg]
\end{equation}
The window function is applied to each analysis frame by computing the element-wise product of the Hann window array, $H$ and each row of the analysis frames matrix, $A_{i,:}$. Result is an $k \times N$ array, $\widetilde{A}$:
\begin{equation}
\label{eqn-WindowMatrix}
\widetilde{A_i} = A_i \odot H
\end{equation}

\paragraph*{}Finally, we perform a \textit{Discrete Fourier Transform} (DFT) to bring each analysis frame from a time domain into a frequency domain \cite{Olson,Peatross}. The Discrete Fourier Transform is applied by producing an $N \times N$ \textit{transform matrix}, often noted at $\mathbb{W}$. Let $\omega^k = e^{\frac{-2\pi i}{N}k}$, then the DFT matrix for a time-space containing $N$ samples
\begin{equation}
\label{eqn-DFTMatrix}
\mathbb{W} = \frac{1}{\sqrt{N}}
\begin{bmatrix}
1 & 1 & 1 & 1 & \hdots & 1 \\
1 & \omega		& \omega^2 & \omega^3 & \hdots & \omega^{N-1} \\
1 & \omega^2	& \omega^4 & \omega^6 & \hdots & \omega^{2(N-1)} \\
1 & \omega^3	& \omega^6 & \omega^9 & \hdots & \omega^{3(N-1)} \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots \\
1 & \omega^{N-1} & \omega^{2(N-1)} & \omega^{3(N-1)} & \hdots & \omega^{(N-1)^2}
\end{bmatrix}
\end{equation}
Each column of the matrix is a complex sinusoidal oscillating with an integer number of periods within the $N$-sample length window \cite{Short,Peatross}.The DFT is applied by a taking the matrix -product of $\mathbb{W}$ and $\widetilde{A}^T$. The transpose of $\widetilde{A}$ then makes each analysis frame into a column vector, which gives the appropriate dimension for multiplication.

@@

\paragraph*{}Each column of the $S$ matrix is now a single-frame that has been moved into a frequency-space representation, thus there are $k$ columns, just as there were $k$ time-series analysis frames. Given the discretized nature of digital audio, the frequency-space representation is not a continuous function, but rather a column vector, where the frequency has been assigned to one of $N$ bins. Since human hearing extends from $20$ Hz to $20$ kHz, audio files such as music and voice recordings are typically sampled around $44.1$ kHz, to ensure that the full range of audio is held by the recording \cite{Olson,Virtanen}.

\paragraph*{}However, standard musical instruments seldom extend above $8$ kHz \cite{Olson,Virtanen,White}. This means that when constructing the spectrogram, we will rarely ever see energy present above this frequency at any time, and the $S$ matrix will contain mostly zero, or zero-like entries. As a result of this, we can simply crop each matrix to only represent frequencies that are only between $0$ Hz and $8000$ Hz. This makes the input array smaller, and eliminates redundant and non-useful information. The number rows in the $S$ matrix is reduced from $N$ down to $N'$. Since each raw audio file may contain a different number of samples to begin with, a different number of analysis frames are created. To ensure all samples are a homogeneous size, we assert that each matrix must have exactly $k'$ columns. If $k' < k$, columns of zeros are added to pad the input, and if $k' > k$, columns are removed.

\paragraph*{}Each spectrogram is now $N' \times k' \times 1$ ($1$ representing a single gray-scale pixel channel, as opposed to $3$ channels for RGB inputs) and effectively encodes the energy distribution of the waveform a a function of both time and frequency. The spectrogram is the first feature used in this model. For this classifier, we have chosen $N' = 560$ and $k' = 256$. For training, a batch of $b$ samples are concatenated into a single array object. \textcolor{blue}{See Neural Network section for more details}. For a batch of $b$ samples of $N' \times k' \times 1$ spectrograms, we shape $X_1$ such that:
\begin{equation}
\label{eqn-X1 shape}
X_1 = \big\{ S^{(0)},S^{(1)},S^{(2)}, ... , S^{(b-1)} \big\} \in \mathbb{R}^{(b \times N' \times k' \times 1)}
\end{equation}
Which is consistent with the shape of the $X_1$ matrix outlined in eqn(\ref{eqn-X1Shape}). This matrix is presented to the \textit{Convolution} branch of the neural network for processing. \textcolor{blue}{See Neural Network section for more details}

>>>>

\paragraph*{}Tuomas Virtanen, machine learning and audio engineer writes in his book, "Computational Analysis of Sound Scene and Events" about specifically how features must be chosen \cite{Virtanen}:
\begin{quote}
For recognition algorithms, the necessary property of the acoustic features is low variability among features extracted from examples assigned to the same class, and at the same time high variability allowing distinction between features extracted from examples assigned to different classes.
\end{quote}
In constructing a neural network classifier, the development of appropriate features is of the utmost importance. To ensure the construction of a suitable model, we derive features based from three sources (i) a spectrogram matrix of the waveform, (ii) the time-space representation of the waveform, and (iii) the frequency-space representation of the waveform. It is important to note that although this algorithm will classify sound waves to instruments, the model will never actually be presented with a waveform directly, instead it will rely solely these features.

\paragraph*{}Once we produce an sufficient set of features, we concatenate them into a single object, $\hat{x}$, called the \textit{feature-vector} \cite{Goodfellow}. In the training process this object, along with the appropriate classification label, $y$ is presented to the neural network for processing. This process of constructing a feature vector from any data set is vital and is used to represent the data set in a far more compact and non-redundant format \cite{Virtanen,Liu}

@@

\paragraph*{}The $21$ features are from each file are organized into one of two array structures. each called a \textit{design matrix}. A design matrix is an N-Dimensional array of real floating-point numbers that stores the features in a convenient way for the neural network to digest. For this hybrid neural-network, we use two matrices, $X_1$ and $X_2$, each containing data from $b$ audio file samples. each has shape given by:
\begin{equation}
\label{eqn-X1Shape}
X_1 \in \mathbb{R}^{(b \times N' \times k \times 1)}
\end{equation}
\begin{equation}
\label{eqn-X2Shape}
X_2 \in \mathbb{R}^{(b \times P)}
\end{equation}
Where $Q' \times R' \times 1$ gives the shape of a single spectrogram matrix $S$, described in sect. (\ref{subsec-spectrogram}). Similarly, $P$ is the number of features derived from the time and frequency-representation of features outlined in sec. (\ref{subsec-time}) and sec.(\ref{subsec-frequency}).The first axis, with size $b$ indicates that there are $b$ samples stored in that design matrix.
A row of $X_2$ has shape $1 \times P$ where there are $P$ features within the \textit{feature-vector} for each sample. Matrix $X_1$ is presented to the convolutional branch of the network to be process as an a collection of images, and matrix $X_2$ is presented to the perceptron branch to be processed as a collection of vectors.

@@

\paragraph*{}The field of neural classification is well studied in the application of image-processing. Many large-scale, and even introductory projects find themselves under the umbrella of image classification \cite{Geron,Goodfellow,Loy,Mierswa}. As a result, model architectures for image processing related tasks are well-explored and have shown experimentally successful behavior. Following this success, it make senses to provide an image-like representation of a sound wave as a feature. We do this in the form of a spectrogram matrix.

\paragraph*{}A spectrogram is a representation of the energy distribution of a sound wave as a function of both frequency and time. In a conventional spectrogram, the passing of time is shown along the $x$-axis, and the frequency spectrum is shown on the $y$-axis. Thus each point in the 2-Dimensional space is an energy at a given time and frequency. Examples spectrograms from the wave form data set are shown in Fig. (\ref{fig-spectrograms}).

@@

\paragraph*{}A spectrogram is produced by the method of \textit{frame-blocking}, which is very prevalent in audio signal classification. Frame-blocking takes a raw waveform or signal, $s$ and decomposes it into a set of analysis frames, $a_i$, with each being $N$ samples in length, and has a fixed overlap with the next adjacent frame. Each of the $k$ frames then allows for a section of the signal in somewhat stationary state \cite{Liu,Zhang,Kahn,Virtanen}. For this project, we have chosen to use frames of size $N = 4096$ with a $75\%$ or $3072$ sample overlap. We also choose to use exactly $k = 256$ frames. Waveforms that are too short are padded with zeros, and waveforms that are too long are truncated. The audio has been sampled at $f_s = 44100$ samples/second, so ach frame represents a slice of time that is about $0.1$ seconds long.

@@

\paragraph*{}After frame-blocking, we apply a \textit{windowing function} to each frame. A standard \textit{Hann Window} of $N$ samples is generated as a $1 \times N$ row-array, $H$. The $n$-th index in a Hann window with $N$ samples is defined:
\begin{equation}
\label{eqn-Hann}
H[n] = \frac{1}{2}\bigg[ 1 - \cos\Big( \frac{2\pi n}{N-1}\Big)\bigg]
\end{equation}
This window function is applied to each analysis frame by computing the element-wise product of the Hann window array, $H$ and each row of the analysis frames matrix, $A_{i,:}$. Result is an $k \times N$ array, $\widetilde{A}$:
\begin{equation}
\label{eqn-WindowMatrix}
\widetilde{A_i} = A_i \odot H
\end{equation}
The window function allows for a cleaner transform into frequency space \cite{Virtanen}.

\paragraph*{}Finally, we perform a \textit{Discrete Fourier Transform} (DFT) to bring each analysis frame from a time domain into a frequency domain \cite{Olson,Peatross}. The Discrete Fourier Transform is applied by producing an $N \times N$ \textit{transform matrix}, often noted at $\mathbb{W}$. Let $\omega^k = e^{\frac{-2\pi i}{N}k}$, then the DFT matrix for a time-space containing $N$ samples
\begin{equation}
\label{eqn-DFTMatrix}
\mathbb{W} = \frac{1}{\sqrt{N}}
\begin{bmatrix}
1 & 1 & 1 & 1 & \hdots & 1 \\
1 & \omega		& \omega^2 & \omega^3 & \hdots & \omega^{N-1} \\
1 & \omega^2	& \omega^4 & \omega^6 & \hdots & \omega^{2(N-1)} \\
1 & \omega^3	& \omega^6 & \omega^9 & \hdots & \omega^{3(N-1)} \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots \\
1 & \omega^{N-1} & \omega^{2(N-1)} & \omega^{3(N-1)} & \hdots & \omega^{(N-1)^2}
\end{bmatrix}
\end{equation}
Each column of the matrix is a complex sinusoidal oscillating with an integer number of periods within the $N$-sample length window \cite{Short,Peatross}.The DFT is applied by a taking the matrix -product of $\mathbb{W}$ and $\widetilde{A}^T$. The transpose of $\widetilde{A}$ then makes each analysis frame into a column vector, which gives the appropriate dimension for multiplication.
\begin{equation}
\label{eqn-DFT}
\text{DFT}\big[ \widetilde{A} \big] = \mathbb{W}\widetilde{A}^T
\end{equation}

@@

\paragraph*{}Each column of the $S$ matrix is now a single-frame that has been moved into a frequency-space representation, thus there are $k$ columns, just as there were $k$ time-series analysis frames. Given the discretized nature of digital audio, the frequency-space representation is not a continuous function, but rather a column vector, where the frequency has been assigned to one of $N$ bins, ranging from $-f_s / 2$ to $+f_s/2$.

\paragraph*{}Standard western musical instruments seldom extend above $6$ kHz \cite{Olson,Virtanen,White}. This means that when constructing the spectrogram, we will rarely ever see energy present above this frequency at any time, and the $S$ matrix will contain mostly zero, or zero-like entries. As a result of this, we can simply crop each matrix to only represent frequencies that are only between $0$ Hz and $6000$ Hz. This makes the input array smaller, and eliminates redundant and non-useful information. The number rows in the $S$ matrix is reduced from $N$ down to $N'$.

\paragraph*{}Each spectrogram is now $N' \times k \times 1$ ($1$ representing a single gray-scale pixel channel, as opposed to $3$ channels for RGB inputs) and effectively encodes the energy distribution of the waveform as a function of both time and frequency. The spectrogram is the first feature used in this model. For this classifier, we have chosen $N' = 558$ and $k = 256$. For training, a batch of $b$ samples are concatenated into a single array object. \textcolor{blue}{See Neural Network section for more details}. For a batch of $b$ samples of $N' \times k \times 1$ spectrograms, we shape $X_1$ such that:
\begin{equation}
\label{eqn-X1 shape}
X_1 = \big\{ S^{(0)},S^{(1)},S^{(2)}, ... , S^{(b-1)} \big\} \in \mathbb{R}^{(b \times N' \times k \times 1)}
\end{equation}
Which is consistent with the shape of the $X_1$ matrix outlined in eqn(\ref{eqn-X1Shape}). This matrix is presented to the \textit{Convolution} branch of the neural network for processing. \textcolor{blue}{See Neural Network section for more details}
